---
title: "Basic Protocol 3"
author: "Wenyan Li"
date: "2023-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Support script for basic protocol 3

```{r}
# load required packages
library(rstudioapi)
library(Seurat)
library(patchwork)
library(S4Vectors)
library(tidyverse)
library(ggplot2)

# set a seed to re-produce pseudorandom numbers
set.seed(99)

# Obtain the path of this rmarkdown file and assign it to object "get_path"
get_path <- dirname(rstudioapi::getSourceEditorContext()$path)

# Set "get_path" as Working Directory
setwd(get_path)

# Save images and data to this folder
saveTo <- "Protocol 3 output"


```


```{r}

# Create a function to save output images

func_save_images <- function(image.object, # plot objects
                             image.name, # name of the plot
                             image.path, # path to save image
                             h = 8, # image height in inch
                             w = 15, # image width in inch
                             r = 80, # image resolution
                             isHeatmap = F # check if image is a Seurat heatmap
		){
  
  ifelse(!dir.exists(image.path), 
         dir.create(image.path,
                    recursive = T), 
         FALSE)
  
  if(isHeatmap){
    
      for(i in seq_along(image.name)){
        temp.path <- paste(image.path,
			      image.name[i], 
                           sep = "/")
        
        tiff(filename = paste(temp.path,
                              "tiff",
                              sep = "."),
             width = w, 
             height = h,
             res = r, 
             units = "in")
        
        image.object[[i]]
        dev.off()
      } # end of for 1st loop
    
  }else{
    
      for(i in seq_along(image.name)){
        temp.path <- paste(image.path,
                           image.name[i], 
                           sep = "/")
        
        ggsave(filename = paste(temp.path,
                                "tiff",
                                sep = "."), 
               plot = image.object[[i]],
               width = w, 
               height = h, 
               units = "in", 
               dpi = r, 
               limitsize = FALSE)
        
      } # end of for 2nd loop
    
  } # end of if isHeatmap statement
  
}# end of func_save_images function


```


# 3.1 Load and read data

```{r}
# Choose a folder/experiment to analyse
exp1 <- "BD-Demo-7Bridges-WTA_AbSeq_SMK"
exp2 <- "BD-Demo-RhapTCRBCRdemo"

# Load the Seurat objects as demo_seurat_1 and demo_seurat_2
demo_seurat_1 <- readRDS(list.files(exp1, full.names = T))
demo_seurat_2 <- readRDS(list.files(exp2, full.names = T))

```

# 3.2 Results: Load and read data

```{r}
demo_seurat_1
```

```{r}
demo_seurat_2
```
# 3.3 Curate the Seurat objects

This step can be skipped if AbSeq is not applicable.

```{r}
# Curate the Seurat objects
# get AbSeq list
AbSeq_list_1 <- rownames(demo_seurat_1)[grepl("pAbO", rownames(demo_seurat_1))]
AbSeq_list_2 <- rownames(demo_seurat_2)[grepl("pAbO", rownames(demo_seurat_2))]

# get RNA gene list
RNA_list_1 <- rownames(demo_seurat_1)[!grepl("pAbO", rownames(demo_seurat_1))]
RNA_list_2 <- rownames(demo_seurat_2)[!grepl("pAbO", rownames(demo_seurat_2))]

# subset demo_seurat_1 and demo_seurat_2 by AbSeq and RNA gene lists
demo_AbSeq_1 <- subset(demo_seurat_1,features=AbSeq_list_1)
demo_AbSeq_2 <- subset(demo_seurat_2,features=AbSeq_list_2)

demo_seurat_1 <- subset(demo_seurat_1,features=RNA_list_1)
demo_seurat_2 <- subset(demo_seurat_2,features=RNA_list_2)

# create new Seurat objects
demo_seurat_1@assays[["AB"]] <- GetAssay(demo_AbSeq_1,assay = "RNA")
demo_seurat_2@assays[["AB"]] <- GetAssay(demo_AbSeq_2,assay = "RNA")

# change default assay to AB
DefaultAssay(demo_seurat_1) <- "AB"
demo_seurat_1$nCount_AB <- colSums(x = demo_seurat_1, slot = "counts")  # nCount_RNA
demo_seurat_1$nFeature_AB <-  colSums(x = GetAssayData(object = demo_seurat_1, slot = "counts") > 0)

DefaultAssay(demo_seurat_2) <- "AB"
demo_seurat_2$nCount_AB <- colSums(x = demo_seurat_2, slot = "counts")  # nCount_RNA
demo_seurat_2$nFeature_AB <-  colSums(x = GetAssayData(object = demo_seurat_2, slot = "counts") > 0)

# remove demo_AbSeq_1 and demo_AbSeq_2
remove(demo_AbSeq_1, demo_AbSeq_2)
```

For this protocol, the 'Sample_Name' cell annotation is duplicated as 'smk'. 
This step can be skipped if sample tagging is not applicable.

```{r}
demo_seurat_1$smk <- demo_seurat_1$Sample_Name
demo_seurat_2$smk <- demo_seurat_2$Sample_Name
```


# 3.4 Results: Curate the Seurat objects

```{r}
demo_seurat_1
```

```{r}
demo_seurat_2
```

# 3.5 Data normalisation, scaling and finding clusters

```{r}

# Build function

# The size of AbSeq panels varies, reduce the number 
# for ab_pc_num if your panel is smaller than 10.
func_quick_process <- function(demo_seurat,
                               ab_pc_num = 10, # number of PCA components to use for protein
                               rna_pc_num = 15, # number of PCA components to use for RNA
                               ab_reduction_res = 0.8, # cluster resolution for protein
                               rna_reduction_res = 0.8) # cluster resolution for RNA
  
     {
  
      # check if the Seurat object has protein assay
      if("AB" %in% Seurat::Assays(demo_seurat))
        
          {  
           Seurat::DefaultAssay(demo_seurat) <- 'AB'
  
           # Normalize and scale data
           demo_seurat <- demo_seurat %>% 
                          Seurat::NormalizeData()  %>% 
                          Seurat::FindVariableFeatures() %>% 
                          Seurat::ScaleData() 

	  # perform PCA
           demo_seurat <- Seurat::RunPCA(object = demo_seurat, 
                                         reduction.name = 'apca')

	  # perform UMAP
           demo_seurat <- Seurat::RunUMAP(demo_seurat, 
                                          reduction = 'apca', 
                                          dims = 1:ab_pc_num, 
                                          assay = 'AB', 
                                          reduction.name = 'adt.umap',
			               reduction.key = 'adtUMAP_')

	  # Find clusters
           demo_seurat <- Seurat::FindNeighbors(demo_seurat, 
                                                reduction = "apca", 
                                                dims = 1:ab_pc_num)

           demo_seurat <- Seurat::FindClusters(demo_seurat, 
                                               resolution = ab_reduction_res, 
                                               graph.name = "AB_snn")

           } # end of if statement
  
     # Change default assay to RNA
     Seurat::DefaultAssay(demo_seurat) <- "RNA"

     # Calculate percentages of mitochondrial gene expression for every cell. If it is a
     #  targeted sequencing, then the output of this commend line is all zeros.
     demo_seurat <- Seurat::PercentageFeatureSet(demo_seurat, 
                                                 pattern = "^MT[-|.]", 
                                                 col.name = "percent.mt") 

     # find top most variant genes
     demo_seurat <- demo_seurat %>% 
                    Seurat::NormalizeData() %>% 
	           Seurat::FindVariableFeatures(., 
		                               selection.method = "vst")

     # scale data
     demo_seurat <- Seurat::ScaleData(demo_seurat, 
                                      verbose = FALSE)

     # perform PCA
     demo_seurat <- Seurat::RunPCA(demo_seurat, 
                                   npcs = rna_pc_num, 
                                   verbose = FALSE)

     # perform UMAP
     demo_seurat <- Seurat::RunUMAP(demo_seurat, 
                                    reduction = "pca", 
                                    dims = 1:rna_pc_num)

     # Find clusters
     demo_seurat <- Seurat::FindNeighbors(demo_seurat, 
                                          reduction = "pca", 
                                          dims = 1:rna_pc_num)

     demo_seurat <- Seurat::FindClusters(demo_seurat, 
                                         resolution = rna_reduction_res)

     demo_seurat <- Seurat::BuildClusterTree(demo_seurat)

     # return Seurat object as output
     return(demo_seurat)

} # end of func_quick_process function

# Use the above function to process the Seurat objects
demo_seurat_1 <- func_quick_process(demo_seurat_1)

demo_seurat_2 <- func_quick_process(demo_seurat_2)


```


Save raw Seurat object as R data

```{r}
# Create a folder called saveTo
ifelse(!dir.exists(saveTo), 
       dir.create(saveTo, 
                  recursive = T), 
       FALSE)

# Save Seurat objects to local drive
save(demo_seurat_1,
     demo_seurat_2,
     file = paste(saveTo, 
                  "raw demo Seurat objects.RData", 
                  sep = "/")
    )


```



# 3.6 Results: Data normalization, scaling and finding clusters

```{r}
# QC plots – check mitochondrial gene percentages
p1 <- Seurat::VlnPlot(demo_seurat_1, 
                      features = "percent.mt", 
              group.by = "seurat_clusters") + 
              Seurat::NoLegend() + 
              ggtitle("demo exp 1 - MT gene%")

p2 <- Seurat::VlnPlot(demo_seurat_2, 
                      features = "percent.mt", 
              group.by = "seurat_clusters") + 
              Seurat::NoLegend() + 
              ggtitle("demo exp 2 - MT gene%")

patchwork::wrap_plots(p1 + p2)

func_save_images(image.object = list(patchwork::wrap_plots(p1 + p2)),
                 image.name = "Initial MT genes percentages",
                 image.path = saveTo,
                 h = 5, 
                 w = 5, 
                 r = 300, 
                 isHeatmap = F
                 )


```



```{r}
p3 <- Seurat::FeatureScatter(demo_seurat_1, 
                             feature1 = "nCount_RNA", 
                             feature2 = "nFeature_RNA", 
                             group.by = "seurat_clusters") + 
              scale_x_log10() +
              scale_y_log10() +
              ggtitle("demo exp 1")

p4 <- Seurat::FeatureScatter(demo_seurat_2, 
                             feature1 = "nCount_RNA", 
                             feature2 = "nFeature_RNA", 
                             group.by = "seurat_clusters") + 
              scale_x_log10() +
              scale_y_log10() +
              ggtitle("demo exp 2")

patchwork::wrap_plots(p3 + p4)

func_save_images(image.object = list(patchwork::wrap_plots(p3 + p4)),
                 image.name = "Initial nCounts and nFeatures",
                 image.path = saveTo,
                 h = 5, 
                 w = 11, 
                 r = 300, 
                 isHeatmap = F
                 )


```



```{r}
# filter out cells with MT genes percentage > 50 (%) and 
# cells with low nFeature_RNA
subset_demo_seurat_1 <- subset(demo_seurat_1, 
                               subset = percent.mt < 50 & 
                                        nFeature_RNA > 200, 
                               invert = F)

subset_demo_seurat_2 <- subset(demo_seurat_2, 
                               subset = percent.mt < 50 & 
                                        nFeature_RNA > 100, 
                               invert = F)

p5 <- Seurat::VlnPlot(subset_demo_seurat_1, 
                      features = "percent.mt", 
                      group.by = "seurat_clusters") + 
              Seurat::NoLegend() + 
              ggtitle("demo exp 1 - MT gene%")

p6 <- Seurat::VlnPlot(subset_demo_seurat_2, 
                      features = "percent.mt", 
                      group.by = "seurat_clusters") + 
              Seurat::NoLegend() + 
              ggtitle("demo exp 2 - MT gene%")

patchwork::wrap_plots(p5 + p6)

func_save_images(image.object = list(patchwork::wrap_plots(p5 + p6)),
                 image.name = "Subset - MT genes percentages (before re-clustering)",
                 image.path = saveTo,
                 h = 5, 
                 w = 5, 
                 r = 300, 
                 isHeatmap = F
                 )

p7 <- Seurat::FeatureScatter(subset_demo_seurat_1, 
                             feature1 = "nCount_RNA", 
                             feature2 = "nFeature_RNA", 
                             group.by = "seurat_clusters") + 
              scale_x_log10() +
              scale_y_log10() +
              ggtitle("demo exp 1")

p8 <- Seurat::FeatureScatter(subset_demo_seurat_2, 
                             feature1 = "nCount_RNA", 
                             feature2 = "nFeature_RNA", 
                             group.by = "seurat_clusters") + 
              scale_x_log10() +
              scale_y_log10() +
              ggtitle("demo exp 2")

patchwork::wrap_plots(p7 + p8)

func_save_images(image.object = list(patchwork::wrap_plots(p7 + p8)),
                 image.name = "Subset - nCounts and nFeatures (before re-clustering)",
                 image.path = saveTo,
                 h = 5, 
                 w = 11, 
                 r = 300, 
                 isHeatmap = F
                 )


```



Re-cluster subsetted samples
```{r}

subset_demo_seurat_1 <- func_quick_process(subset_demo_seurat_1)

subset_demo_seurat_2 <- func_quick_process(subset_demo_seurat_2)

```



```{r}
# UMAP plots

p9 <- Seurat::DimPlot(subset_demo_seurat_1, 
                      reduction = "umap", 
                      group.by = "seurat_clusters") + 
              ggtitle("demo exp 1")

p10 <- Seurat::DimPlot(subset_demo_seurat_2, 
                       reduction = "umap", 
                       group.by = "seurat_clusters") + 
              ggtitle("demo exp 2")

patchwork::wrap_plots(p9+p10)

func_save_images(image.object = list(patchwork::wrap_plots(p9 + p10)),
                 image.name = "Subset - UMAP group by clusters",
                 image.path = saveTo,
                 h = 5, 
                 w = 11, 
                 r = 300, 
                 isHeatmap = F
                 )

p11 <- Seurat::DimPlot(subset_demo_seurat_1, 
                       reduction = "umap", 
                       group.by = "smk") + 
               ggtitle("demo exp 1")

p12 <- Seurat::DimPlot(subset_demo_seurat_2, 
                       reduction = "umap", 
                       group.by = "smk")  + 
               ggtitle("demo exp 2")

patchwork::wrap_plots(p11+p12)

func_save_images(image.object = list(patchwork::wrap_plots(p11 + p12)),
                 image.name = "Subset - UMAP group by sample tag",
                 image.path = saveTo,
                 h = 5, 
                 w = 13, 
                 r = 300, 
                 isHeatmap = F
                 )


```



Save subsetted Seurat objects
```{r}
# Save Seurat objects to local drive
save(subset_demo_seurat_1,
     subset_demo_seurat_2,
     file = paste(saveTo, 
                  "subsetted demo Seurat objects.RData", 
                  sep = "/"))

```




